An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network

### 2025-11-30 paper/main.tex
weiter

\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage[hidelinks]{hyperref}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{caption}
\usepackage{csquotes}
\usepackage{float}
\usepackage{derivative}
\usepackage{subcaption}
\usepackage{bbm}

@@

\begin{abstract}
	\noindent
	We study how the numerical precision of weights, activations, and gradients in
	neural networks can be reduced without sacrificing too much of the performance.
	Using MNIST as a general dataset (and FashionMNIST for the decimal rounding),
	we benchmark uniform quantization (mapping real values to evenly spaced levels
	via a single step size), mixed-precision formats, and post-training decimal
	rounding. We evaluate precision settings (FP64/FP32/FP16) across all tensors
	during training, experiment with an inference-only quantization path on
	FP64-trained models, and test automatic mixed precision with gradient clipping.
	Metrics include test accuracy, training/inference time, and parameter storage.
	We compare end-to-end low-precision training, inference-only quantization, and
	mixed-precision with gradient clipping, and evaluate their impact on test
	accuracy and efficiency. The results on the used test setup show that (i)
	16-bit end-to-end training does not significantly decrease test accuracy for
	the MNIST dataset, (ii) three-decimal-place weight rounding while keeping
	32-bit precision incurs minimal accuracy loss. These findings verify limits for
	memory- and energy-constrained usecases like inference in mobile devices.
\end{abstract}

@@

Researchers have used a range of techniques, like stochastic gradient descent

@@

lead to significant improvements in memory usage and processing speed. Even
when precision is reduced to very low levels, models can still perform
competitively on complex tasks. However, while these early results are
promising, the overall effects of lowering precision, especially how it
influences model accuracy, training behavior, and storage efficiency, are still
not fully understood and deserve more detailed investigation. \\\\ This is

@@

\noindent
DoReFa-Net extends this concept by quantizing \textit{weights},
\textit{activations}, and \textit{gradients} to low-precision
formats~\cite{zhou2018dorefa}. The method allows using \textit{arbitrary
	bitwidths}, i.e., the number of bits is not restricted to standard IEEE
floating-point formats, but can be any small integer (for example, 1, 2, or 4
bits for weights and slightly higher bitwidths for activations and gradients).
This is flexible and makes it possible to set the precision specific to
hardware constraints or application requirements, but the training of the
network is still end-to-end.\\

@@

\subsubsection*{Post-Training Quantization}
Post-training quantization begins with an already trained full-precision model
and converts it to a lower precision without (or with minimal) additional
training. An often used approach is integer-only quantization with symmetric
scaling~\cite{jacob2018quantization}. Here, both weights and activations are
mapped to signed integers (for example 8-bit) using a scale factor that is the
same across positive and negative values (therefore symmetric). This design
allows the entire inference flow to be implemented with integer arithmetic,
which is better for embedded or mobile hardware.\\

\noindent
Methods such as GPTQ (Gradient Post-Training Quantization) have been tested to
quantize large transformer models. Examples are GPT language models with
minimal or no retraining~\cite{frantar2022gptq}. GPTQ treats quantization as an
optimization problem by using approximations of the local curvature to choose
quantization levels that minimize the error introduced by replacing
full-precision weights with low-precision ones. That makes this approach better
for very large models where a full retraining with quantization would be too
expensive.\\

\noindent
SmoothQuant focuses more on activations~\cite{xiao2024smoothquant}. It rescales
activations and compensates for this rescaling in the weight matrices so that
the activation distributions become more smooth and better aligned with the
range supported by low-precision integer formats. This improves the robustness
of post-training quantization, especially in transformer architectures where
activation outliers can otherwise complicate the scaling and lead to large
quantization errors.

\subsection{Artificial Neural Networks}
A typical neuron in a neural network performs a weighted sum of its inputs,
followed by an activation function which results in the neuron output $h$
depicted in~\eqref{eq:activation}. Here, $\omega\in\mathbb{R}^n$ is the weight
vector, $b \in \mathbb{R}$ is the bias, $x \in \mathbb{R}^n$ is the input, and
$\sigma$ is a nonlinear activation function (e.g., sigmoid or ReLU).

@@

\noindent
In our MLP, hidden layers use the Rectified Linear Unit (ReLU), so
$\sigma=\mathrm{ReLU}$, while the output layer is linear, producing raw scores
(also called logits) $z$ later used in~\eqref{eq:ce}. ReLU~\eqref{eq:relu}
keeps positive activations linear and sets negatives to zero. An additional
benefit is that the computation is very fast and simple compared to more
complex activation functions.

@@

with $W$ and $b$ the weight matrix and bias of the last linear layer, $h$ the
layer input, and $e_y$ the one-hot vector of the target class. Here,
$\mathbbm{1}\{k=y\}$ is the indicator: it equals $1$ if $k=y$ and $0$
otherwise.\\

\noindent
Memory and computational requirements during training are substantial, as
weights, gradients, activations, and parameters must be stored and updated
iteratively. Training and inference can both be optimized using precision
reduction and related techniques, but inference is more well suited to
optimizations because it only requires a forward pass.

@@

\section{Experimental Design}
The effect of numerical precision on training and inference is evaluated
through five experiment groups run on standard image classification tasks. The
main dataset MNIST\footnote{LeCun, Cortes, and Burges:
	\url{http://yann.lecun.com/exdb/mnist/}.} ($28\times 28$ grayscale images of
handwritten digits, train/test split $60.000$/$10.000$, 10 classes, examples
shown in Figure~\ref{fig:mnistSample}) is used to determine precision effects
on accuracy and runtime under a simple workload, while an after-training
rounding experiment is performed on FashionMNIST\footnote{Xiao, Rasul, and
	Vollgraf: \url{https://github.com/zalandoresearch/fashion-mnist}.} ($28\times
28$ grayscale images of clothing items, train/test split $60.000$/$10.000$, 10
classes) so the impact of using coarse value sets is better observable.\\

@@

The classifier used is a fully connected network
$784\rightarrow128\rightarrow64\rightarrow10$ with ReLU between linear layers
(Figure~\ref{fig:mlpStructure}). Inputs are normalized with mean $0.1307$ and
std $0.3081$. Unless stated otherwise, we train with cross-entropy loss and SGD
(learning rate $0.01$, momentum $0.9$), batch size $64$, and on CUDA.\\

@@

For the MNIST experiments (i) end-to-end precision, (ii) timing, (iii)
inference-only precision reduction, and (iv) automatic mixed precision (AMP)
with gradient clipping, we use 5-fold cross-validation on the training set with
random equal splits. Unless noted otherwise, figures report fold means, and no
test set tuning is performed. In the end-to-end runs, every floating point
tensor appearing inside the training loop (parameters, activations, gradients,
and optimizer state) is kept at the target precision during training and
evaluation. The timing experiment reuses these runs to measure mean per-epoch
training time and mean evaluation time. For inference-only reduction, we first
train a baseline model in FP64 and then do inference on copies cast to FP32 and
FP16 without any further optimization (so no \textit{retraining} on the lower
precision).\\

The AMP experiment is run on MNIST with gradient clipping, comparing an AMP
setup (FP16 computations with FP32 master weights) against a pure FP32
baseline. We try different global-norm and per-element value-clipping
thresholds to check stability and report test accuracy and training/inference
times. We also run a one-off rounding experiment on FashionMNIST (batch size
$256$, FP32 training): after training, only the stored weights are rounded to
$k$ decimal places ($k{=}1\ldots14$) and inference remains FP32. Inputs and
intermediate computations stay at full precision. We report test accuracy as a
function of $k$ and note the impact on storage.

\section{Results}
In the first experiment, every floating-point tensor that is used inside the
training loop was converted to a lower precision data type. This therefore
includes the weights and biases, activations, gradients, optimizer states, etc.
The types tried were FP64, FP32, and FP16.

@@

First step is using global norm clipping, all gradients are collectively scaled
to ensure their combined norm does not exceed a specified threshold, promoting
stable updates across the network. In second step per-element clipping
constrains each individual gradient component within a fixed range, directly
limiting extreme values. These methods are evaluated under Automatic Mixed
Precision (AMP) to compare their impact on training stability, model accuracy,
and computation time across different precision types.

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage[hidelinks]{hyperref}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{caption}
\usepackage{csquotes}
\usepackage{float}
\usepackage{derivative}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{siunitx}

@@

\begin{abstract}
	\noindent
	We study how the numerical precision of weights, activations, and gradients in
	neural networks can be reduced without sacrificing too much of the performance.
	Using MNIST as a general dataset, and FashionMNIST for the decimal rounding, we
	benchmark training with reduced precision (FP64/FP32/FP16), mixed-precision
	formats, and post-training decimal rounding. We evaluate precision settings
	across all tensors during training, experiment with an inference-only casting
	to lower-precision floating point on FP64-trained models, and test Automatic
	Mixed Precision (AMP) with gradient clipping. Metrics include test accuracy,
	training/inference time, and parameter storage. We compare end-to-end
	low-precision training, inference-only quantization, and mixed-precision with
	gradient clipping, and evaluate their impact on test accuracy and efficiency.
	The results on the used test setup show that (i) 16-bit end-to-end training
	does not significantly decrease test accuracy for the MNIST dataset, (ii)
	three-decimal-place weight rounding while keeping 32-bit precision incurs
	minimal accuracy loss. These findings verify limits for memory- and
	energy-constrained usecases like inference in mobile devices.
\end{abstract}

@@

Researchers have used a range of techniques, like Stochastic Gradient Descent

@@

lead to significant improvements in memory usage and inference efficiency
without degrading the accuracy too much. For example, integer-only INT8
quantization~\cite{jacob2018quantization} and also binary-weight and
binary-activation CNNs evaluated on ImageNet~\cite{rastegariECCV16}. However,
while these early results are promising, the accuracy/efficiency trade-off
depends on architecture, task, storage efficiency, and training setup, so
precision reduction still deserves more detailed investigation. \\\\ This is

@@

\noindent
DoReFa-Net extends this concept by quantizing \textit{weights},
\textit{activations}, and \textit{gradients} to low-precision
formats~\cite{zhou2018dorefa}. The method allows using \textit{arbitrary
	bitwidths} which means the number of bits is not restricted to standard IEEE
floating-point formats, but can be any small integer (for example, 1, 2, or 4
bits for weights and slightly higher bitwidths for activations and gradients).
This is flexible and makes it possible to set the precision specific to
hardware constraints or application requirements, but the training of the
network is still end-to-end.\\

@@

\subsubsection*{Post-Training Quantization}
Post-training quantization begins with an already trained full-precision model
and converts it to a lower precision without (or with minimal) additional
training. An often used approach is integer quantization with symmetric
scaling~\cite{jacob2018quantization}. Here, both weights and activations are
mapped to signed integers (for example 8-bit) using a scale factor that is the
same across positive and negative values (therefore symmetric). This design
allows the entire inference flow to be implemented with integer arithmetic,
which is better for embedded or mobile hardware.\\

\noindent
Methods such as GPTQ (Gradient Post-Training Quantization) have been tested to
quantize large transformer models. Examples are GPT language models with
minimal or no retraining~\cite{frantar2022gptq}. GPTQ handles quantization as
an optimization problem by using approximations of the local curvature to
choose quantization levels that minimize the error introduced by replacing
full-precision weights with low-precision ones. That makes this approach better
for very large models where a full retraining with quantization would be too
expensive.\\

\noindent
SmoothQuant focuses more on activations~\cite{xiao2024smoothquant}. It rescales
activations and compensates for this rescaling in the weight matrices so that
the activation distributions become more smooth and better aligned with the
range supported by low-precision integer formats. This improves the
post-training quantization, especially in transformer architectures where
activation outliers can otherwise complicate the scaling and lead to large
quantization errors.

\subsection{Artificial Neural Networks}
A typical neuron in a neural network performs a weighted sum of its inputs,
followed by an activation function which results in the neuron output $h$
depicted in~\eqref{eq:activation}. Here, $\omega\in\mathbb{R}^n$ is the weight
vector, $b \in \mathbb{R}$ is the bias, $x \in \mathbb{R}^n$ is the input, and
$\sigma$ is a nonlinear activation function (sigmoid or ReLU, for example).

@@

\noindent
In our MLP, hidden layers use the Rectified Linear Unit (ReLU), so
$\sigma=\mathrm{ReLU}$, while the output layer is linear, producing raw scores
(also called logits) $z$ later used in~\eqref{eq:ce}. ReLU~\eqref{eq:relu}
keeps positive activations linear and sets negatives to zero. An additional
benefit is that the calculation is very fast and simple compared to more
complex activation functions.

@@

with $W$ and $b$ the weight matrix and bias of the last linear layer, $h$ the
layer input, and $e_y$ the one-hot vector of the target class. Here,
$\mathbbm{1}\{k=y\}$ is the indicator: it equals \num{1} if $k=y$ and \num{0}
otherwise.\\

\noindent
Memory and computational requirements during training are substantial, as
weights, gradients, and activations must be stored and updated iteratively.
Training and inference can both be optimized using precision reduction and
related techniques, but inference is more well suited to optimizations because
it only requires a forward pass.

@@

\section{Experimental Design}
The effect of numerical precision on training and inference is evaluated
through five experiment groups run on standard image classification tasks. The
main dataset MNIST\footnote{LeCun, Cortes, and Burges:
	\url{http://yann.lecun.com/exdb/mnist/}.} ($28\times 28$ grayscale images of
handwritten digits, train/test split \num{60000}/\num{10000}, \num{10} classes,
examples shown in Figure~\ref{fig:mnistSample}) is used to determine precision
effects on accuracy and runtime under a simple workload, while an
after-training rounding experiment is performed on FashionMNIST\footnote{Xiao,
	Rasul, and Vollgraf: \url{https://github.com/zalandoresearch/fashion-mnist}.}
($28\times 28$ grayscale images of clothing items, train/test split
\num{60000}/\num{10000}, \num{10} classes) so the impact of using coarse value
sets is better observable.\\

@@

\noindent
The classifier used is a fully connected network (layers:
$784\rightarrow128\rightarrow64\rightarrow10$) with ReLU between linear layers
(see structure in Figure~\ref{fig:mlpStructure}). The final linear layer
produces \num{10} logits and during training we use \textit{CrossEntropyLoss}
on the logits, softmax is not applied in the forward pass. Optimization uses
SGD with learning rate \num{0.01} and momentum \num{0.9}, batch size \num{64},
on CUDA.\ Unless noted otherwise we train for a small, fixed number of epochs
per fold. Automatic mixed precision runs use \textit{autocast} with a
\textit{GradScaler}. When enabled, gradients are clipped either by global norm
or per-element value as specified in Section~\ref{sec:results}.

@@

\noindent
For the MNIST experiments (i) end-to-end precision, (ii) timing, (iii)
inference-only precision reduction, and (iv) AMP with gradient clipping, we use
5-fold cross-validation on the training set with random equal splits. Unless
noted otherwise, figures report fold means, and no test set tuning is
performed. In the end-to-end runs, every floating point tensor appearing inside
the training loop (parameters, activations, gradients) is kept at the target
precision during training and evaluation. The timing experiment reuses these
runs to measure mean per-epoch training time and mean evaluation time. For
inference-only reduction, we first train a baseline model in FP64 and then do
inference on copies cast to FP32 and FP16 without any further optimization (so
no \textit{retraining} on the lower precision).\\

The AMP experiment is run on MNIST with gradient clipping, comparing an AMP
setup (FP16 computations with FP32 master weights) against a pure FP32
baseline. We try different global-norm and per-element value-clipping
thresholds to check stability and report test accuracy and training/inference
times. We also run a one-off rounding experiment on FashionMNIST (batch size
\num{256}, FP32 training): after training, only the stored weights are rounded
to $k$ decimal places ($k{=}1\ldots14$) and inference remains FP32. Inputs and
intermediate computations stay at full precision. We report test accuracy as a
function of $k$ and note the impact on storage.

\section{Experiments and Results}\label{sec:results}
In the first experiment, every floating-point tensor that is used inside the
training loop was converted to a lower precision data type. This therefore
includes the weights and biases, activations, gradients, optimizer states, etc.
The types tried were FP64, FP32, and FP16.

@@

First step is using global norm clipping, all gradients are collectively scaled
to ensure their combined norm does not exceed a specified threshold, promoting
stable updates across the network. In second step per-element clipping
constrains each individual gradient component within a fixed range, directly
limiting extreme values. These methods are evaluated under AMP to compare their
impact on training stability, model accuracy, and computation time across
different precision types.
